Numerical Investigation of Viscous Flow Instabilities in Multiphase Heterogeneous Porous Media

### 2018-02-07 Papers/ViscousInstabilities/Manuscript_ViscousInstabilities_KC.tex
updates

\begin{abstract} 
%CAUSE --> EFFECT --> MOTIVATION (P1)
  A critical aspect of multiphase flow in porous media is the displacement efficiency that measures the amount of fluid that can be pushed by another fluid driven by pressure gradient. Migration of contaminants and reservoir waterflooding are typical applications where understanding the dynamics of immiscible fluid displacement helps mitigating water resources contamination and improving hydrocarbons production, respectively. Due to large viscosity ratios, flow instabilities at fluids' interface may arise leading to the formation of fingers -- uneven front with elongation at the outside edge of fluids interface, with strong impact on the displacement efficiency.  
%CAUSE --> EFFECT --> GAPS IN KNOWLEDGE --> AIMS OF THE WORK (P2)
Initial studies of viscous instabilities in Hele-Shaw cells indicated that the development of fingers mostly depends on mobility ratio and capillary number, however heterogeneity of the porous domain may also affect the onset of instabilities. Therefore, the main aim of this work is to numerically investigate formation and growth of viscous fingers in heterogeneous porous media. 
%CAUSE --> EFFECT --> FEW TECHS USED  (P3)
The model used here is based on a novel control volume finite element method (CVFEM) formulation with a family of FE-pairs, \PN[n]{m} and \PNDG[n]{m} specially tailored for Darcean flows. Dynamic mesh adaptivity enables capturing fingers development whilst saving computational overheads. Numerical experiments were performed to investigate the impact of viscosity ratio and heterogeneity on Saffmann-Taylor instabilities.
%CAUSE --> EFFECT -->  MAIN FINDINGS  (P4)
Numerical simulations demonstrated that the heterogeneity of the domain triggers the early-onset formation of fingers under prescribed viscosity ratio conditions. Also, effective numerical capture of growth (in particular, tip-splitting) and coalescence of dendritic finger branching induced by large viscosity ratio largely depends on mesh resolution at the fluids interface.

@@

\medskip 
In this work, a novel CVFEM formulation, previously introduced by \citet{gomes_2017}\citep[see also][]{jackson_2015,salinas2015}, is used to numerically investigate formation and growth of viscous fingers in heterogeneous media. The continuity equation is embedded into the pressure equation to enforce mass conservation whilst ensuring that the force balance is preserved. A hybrid family of \PN[n]{m} triangular and tetrahedral FE pairs is used to discretise velocity and pressure. %, in which velocity is represented by $n$ th-order polynomials that are discontinuous across elements whereas pressure is represented by $m$ th-order polynomials that are continuous across the elements.
A sketch of the \PN[1]{2} FE-pair is shown in Fig.~\ref{fig:fem_cv}, in which velocity is represented by discontinuous and piecewise linear basis functions whereas pressure is interpolated through continuous and piecewise quadratic basis functions. Scalar fields are stored in CV space (Fig.~\ref{fig:fem_elem}). The dual pressure and velocity fields are represented simultaneously (through non-linear projections) in FE and CV spaces.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
\section{Brief Summary of Viscous Instabilities}\label{section:ViscousInstabilities}
The study of viscous flow instabilities (\ie fingering) is particularly important in oil exploration due to heterogeneities (\ie natural fractures, permeability and/or porosity characteristics in different zones) of geological formations. A major problem associated with water-flooding processes is the early water-breakthrough caused by high-permeability layers and unfavorable mobility ratios. Water breakthrough and volumetric sweep efficiency (ratio of volumes between the recovered and the injected fluids) are the main determinants of the productive life of the reservoir \citep{riaz_2004, tavassoli_2015}. During immiscible CO$_{2}$-flooding $\left(\text{\eg in CO}_{2}\text{ enhanced oil recovery, CO}_{2}\text{-EOR, processes}\right)$, density of supercritical CO$_{2}$ is usually lower than hydrocarbons, thus viscous fingering and/or channelling may occur.

\medskip
The efficiency of fluid displacement depends upon the ratio of viscous and capillary forces (or capillary number, $N_{c}$, Eqn.~\ref{eq:capillary_number}). 
%Flow instabilities will occur due to: pressure gradient along the interface of fluids; stress forces acting in opposite directions; mobility differences that may exceed critical limits and; heterogeneities of the domain (layering). 
When viscous force of the injected fluid overcomes the capillary force, hydrodynamic instabilities may occur, resulting in the collapse of the interface between fluids and start forming fingers.

@@

\medskip
More recently, \citet{howison_2000} and \citet{praud_2005} provided a comprehensive description of the mathematical formulation of immiscible two-phase flows in Hele-Shaw cells (also know as Saffman $\&$ Taylor problem). For a Hele-Shaw cell of a given size the development flow only depends on the capillary number. Thus if \textit{$N_{c}$} is too high \citet{saffman_1959}, \citet{homsy_1987} and \citet{tabeling_1987} determined that the flow develops into a single steady-state finger which moves through the cell with constant velocity $U$.

\medskip
 Under the assumption that fluids remain immiscible along an interface, surface tension plays an important role in determining the shape and progress of the fingers~\citep{howison_2000}. During the displacement of a fluid by a less viscous one, the expected uniform front \citep{buckley_1942,sheldon_1959} is perturbed leading to an uneven front with elongations of the outside edge of the fluid interface (Fig.~\ref{fig:simple_case}), often referred as \textit{fingers}. In homogeneous domains, fingers start to develop when the surface tension acting on the interface between the fluids exert an opposite force towards the change of shape of the interface. The interface becomes unstable and collapses, taking a curved shape \citep{homsy_1987, jackson_2017}.

@@

From Eqns.~\ref{eq:MR}-\ref{Eqn:CoreyBrooks2}, it is clear that during fluid displacement, the mobility ratio changes as phase saturation, $S_{w}\left(x_{i},t\right)$ and $S_{nw}\left(x_{i},t\right)$, varies in time and space (Eqn.~\ref{eqn:saturation_eqn}). Therefore, with no lack of generality, the MR can be reduced to the viscosity ratio (VR),
\begin{displaymath}
    \text{VR} = \frc{\mu_{i}}{\mu_{j}},
\end{displaymath}
 that will be used in the parametrisation of the numerical simulations conducted in Section \ref{section:results}. Analysis performed in the following sections will make use of this simplified definition as phase saturation $\left(S_{\alpha}\right)$ is a prognostic field which is calculated along with pressure ($p$) and velocity $\left(\mathbf{u}_{\alpha}\right)$ fields.

\medskip
In immiscible displacements, viscous fingering occurs when the viscosity ratio is greater than unity. As surface tension becomes weak, the interface is stressed and becomes unstable leading to the formation of fingers. At this point it should be mentioned that there are two parameters, the Peclet number (Pe) and mobility ratio (M), that determine the stability characteristics. There are always a few dominant fingers that spread and shield the growth of other fingers. The interface of the main finger collapses and starts splitting into new lobes of fingers. One of these new fingers may eventually outgrow the others and then spreads to occupy an increasingly larger width. In the process, the finger reaches a critical width while the concentration gradient of its front becomes steep as a result of stretching caused by the cross-flow; thus causing the tip of the finger to become unstable and splits again, and the pattern repeats itself. Therefore, surface tension plays an essential dual role, it must be weak enough for the tip front to be unstable, but it is also the physical force causing the spreading and ensuing repeated branching \citep{tan_1988, carvalho_2013}.

>>>>

\begin{abstract} 
%CAUSE --> EFFECT --> MOTIVATION (P1)
  A critical aspect of multiphase flow in porous media is the displacement efficiency that measures the amount of fluid that can be pushed by another fluid driven by pressure gradient. Migration of contaminants and reservoir waterflooding are typical applications where understanding the dynamics of immiscible fluid displacement helps mitigating water resources contamination and improving hydrocarbons production, respectively. Due to large viscosity ratios, flow instabilities at fluids' interface may arise leading to the formation of fingers -- uneven front with elongation at the outside edge of fluids interface, with strong impact on the displacement efficiency.  
%CAUSE --> EFFECT --> GAPS IN KNOWLEDGE --> AIMS OF THE WORK (P2)
Initial studies of viscous instabilities in Hele-Shaw cells indicated that the development of fingers mostly depends on mobility ratio and capillary number, however heterogeneity of the porous domain may also affect the onset of instabilities. Therefore, the main aim of this work is to numerically investigate formation and growth of viscous fingers in heterogeneous porous media. 
%CAUSE --> EFFECT --> FEW TECHS USED  (P3)
The model used here is based on a novel control volume finite element method (CVFEM) formulation with a family of FE-pairs, \PN[n]{m} specially tailored for Darcean flows. Dynamic mesh adaptivity enables capturing fingers development whilst saving computational overheads. Numerical experiments were performed to investigate the impact of viscosity ratio and heterogeneity on Saffmann-Taylor instabilities.
%CAUSE --> EFFECT -->  MAIN FINDINGS  (P4)
Numerical simulations demonstrated that the heterogeneity of the domain triggers the early-onset formation of fingers under prescribed viscosity ratio conditions. Also, effective numerical capture of growth (in particular, tip-splitting) and coalescence of dendritic finger branching induced by large viscosity ratio largely depends on mesh resolution at the fluids interface.

@@

\medskip 
In this work, a novel CVFEM formulation, previously introduced by \citet{gomes_2017} \citep[see also][]{jackson_2015,salinas2015}, is used to numerically investigate formation and growth of viscous fingers in heterogeneous porous media. The continuity equation is embedded into the pressure equation to enforce mass conservation whilst ensuring that the force balance is preserved. A hybrid family of \PN[n]{m} triangular and tetrahedral FE pairs is used to discretise velocity and pressure. %, in which velocity is represented by $n$ th-order polynomials that are discontinuous across elements whereas pressure is represented by $m$ th-order polynomials that are continuous across the elements.
A sketch of the \PN[1]{2} FE-pair is shown in Fig.~\ref{fig:fem_cv}, in which velocity is represented by discontinuous and piecewise linear basis functions whereas pressure is interpolated through continuous and piecewise quadratic basis functions. Scalar fields are stored in CV space (Fig.~\ref{fig:fem_elem}). The dual pressure and velocity fields are represented simultaneously (through non-linear projections) in FE and CV spaces.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
\section{Brief Summary of Viscous Instabilities}\label{section:ViscousInstabilities}
The study of viscous flow instabilities (\ie fingering) is particularly important in oil exploration due to heterogeneities (\ie natural fractures, permeability and/or porosity characteristics in different zones) of geological formations. A major problem associated with water-flooding processes is the early water-breakthrough caused by high-permeability layers and unfavorable mobility ratios. Water breakthrough and volumetric sweep efficiency (ratio of volumes between the recovered and the injected fluids) are the main determinants of the productive life of a reservoir \citep{riaz_2004, tavassoli_2015}. During immiscible CO$_{2}$-flooding $\left(\text{\ie in CO}_{2}\text{ enhanced oil recovery, CO}_{2}\text{-EOR, processes}\right)$, viscosity of supercritical CO$_{2}$ is lower than crude oil, thus viscous fingering and/or channelling may occur.

\medskip
The efficiency of fluid displacement depends upon the ratio of viscous and capillary forces (or capillary number, $N_{c}$, Eqn.~\ref{eq:capillary_number}). 
%Flow instabilities will occur due to: pressure gradient along the interface of fluids; stress forces acting in opposite directions; mobility differences that may exceed critical limits and; heterogeneities of the domain (layering). 
When viscous force of the injected fluid overcomes the capillary force, hydrodynamic instabilities may occur, resulting in the collapse of the interface between fluids and fingers start to form.

@@

\medskip
More recently, \citet{howison_2000} and \citet{praud_2005} provided a comprehensive description of the mathematical formulation of immiscible two-phase flows in Hele-Shaw cells (also know as Saffman $\&$ Taylor problem). For a Hele-Shaw cell of a given size, flow development depends only on the capillary number. Thus if \textit{$N_{c}$} is too high, \citet{saffman_1959}, \citet{homsy_1987} and \citet{tabeling_1987} determined that the flow develops into a single steady-state finger which moves through the cell with constant velocity $U$.

\medskip
 Under the assumption that fluids remain immiscible along an interface, surface tension plays an important role in determining the shape and progress of the fingers~\citep{howison_2000}. During the displacement of a fluid by a less viscous one, the expected uniform front \citep{buckley_1942,sheldon_1959} is perturbed leading to an uneven front with elongations at the outside edge of the fluid interface (Fig.~\ref{fig:simple_case}), often referred as \textit{fingers}. In homogeneous domains, fingers start to develop when the surface tension acting on the interface between the fluids exert an opposite force towards the change of shape of the interface. The interface becomes unstable and collapses, taking a curved shape \citep{homsy_1987, jackson_2017}.

@@

From Eqns.~\ref{eq:MR}-\ref{Eqn:CoreyBrooks2}, it is clear that during fluid displacement, the mobility ratio changes as phase saturation, $S_{w}\left(x_{i},t\right)$ and $S_{nw}\left(x_{i},t\right)$, varies in time and space (Eqn.~\ref{eqn:saturation_eqn}). Therefore, with no lack of generality, the MR can be reduced to the viscosity ratio (VR),
\begin{displaymath}
    \text{VR} = \frc{\mu_{i}}{\mu_{j}},
\end{displaymath}
 that will be used in the parametrisation of the numerical simulations conducted in Section \ref{section:results}. Analysis performed in the following sections will make use of this simplified definition as phase saturation $\left(S_{\alpha}\right)$ is a time- and spatial-dependent prognostic field which is calculated along with pressure ($p$) and velocity $\left(\mathbf{u}_{\alpha}\right)$ fields.

\medskip
In immiscible displacements, viscous fingering occurs when the viscosity ratio is greater than unity. As surface tension becomes weak, the interface is stressed and becomes unstable leading to the formation of fingers. At this point it should be mentioned that there are two parameters -- Peclet number (Pe) and mobility ratio (MR), that determine the stability characteristics. There are always a few dominant fingers that spread and shield the growth of other fingers. The interface of the main finger collapses and starts splitting into new lobes of fingers. One of these new fingers may eventually outgrow the others and then spreads to occupy an increasingly larger width. In the process, the finger reaches a critical width while the saturation %concentration gradient 
of its front becomes steep as a result of stretching caused by the cross-flow; thus causing the tip of the finger to become unstable and splits again, and the pattern repeats itself. Therefore, surface tension plays an essential dual role, it must be weak enough for the tip front to be unstable, but it is also the physical force causing the spreading and ensuing repeated branching \citep{tan_1988, carvalho_2013}.
